---
title: "webinar_code"
author: "Emily Nordmann"
date: "2025-03-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Webinar

This Rmd file contains all the code you need for the webinar. I would encourage you to make notes in this Rmd as we're going along so that everything is all in the one place.

## Set-up

```{r}
# if you get the error message "there is no package called X" it means you need to install it - see the guidance notes
library(tidyverse)
library(jtools)
library(ggExtra)
dat <- read_csv("smarvus_complete_050524.csv")
```


```{r}

demo <- dat %>%
  select(unique_id, age, gender, degree_major)

STARS <- dat %>%
  select(unique_id, Q7.1_1:Q7.1_24) %>% # select the STARS columns
  filter(Q7.1_24 == 1) %>% # remove those who failed the attention check
  select(-Q7.1_24) %>% # remove the attention check column
  pivot_longer(cols = Q7.1_1:Q7.1_23, names_to = "item", values_to = "score") %>% # transform to long-form
  group_by(unique_id) %>% # group by participant
  summarise(stars_score = mean(score, na.rm = TRUE)) %>% # calculate mean STARS score for each ppt
  ungroup() %>% # ungroup so it doesn't mess things up later
  drop_na() #  get rid of NAs so they don't cause us havoc

# Brief Fear of Negative Evaluation Scale

BNFE <- dat %>%
  select(unique_id, Q11.1_1:Q11.1_9) %>% # select the BNFE columns
  filter(Q11.1_9 == 3) %>% # remove those who failed the attention check
  select(-Q11.1_9) %>% # remove the attention check column
  pivot_longer(cols = Q11.1_1:Q11.1_8, names_to = "item", values_to = "score") %>% # transform to long-form
  group_by(unique_id) %>% # group by participant
  summarise(bnfe_score = mean(score, na.rm = TRUE)) %>% # calculate mean BNFE score for each ppt
  ungroup() %>% # ungroup so it doesn't mess things up later
  drop_na() #  get rid of NAs so they don't cause us havoc

# Intolerance of Uncertainty.

IUS <- dat %>%
  select(unique_id, Q14.1_1:Q14.1_12) %>% # select the BNFE columns
  pivot_longer(cols = Q14.1_1:Q14.1_12, names_to = "item", values_to = "score") %>% # transform to long-form
  group_by(unique_id) %>% # group by participant
  summarise(ius_score = mean(score, na.rm = TRUE)) %>% # calculate mean IUS-SF score for each ppt
  ungroup() %>% # ungroup so it doesn't mess things up later
  drop_na() #  get rid of NAs so they don't cause us havoc

# join everything together

dat_joined <- inner_join(x = demo, y = STARS, by = "unique_id") %>%
  inner_join(BNFE, by = "unique_id") %>%
  inner_join(IUS, by = "unique_id")

# convert categorical data to factors

dat_joined <- dat_joined %>%
  mutate(age = as.factor(age),
         gender = as.factor(gender),
         degree_major = as.factor(degree_major))

# take a subset of the data because some groups have very small sample sizes

dat_final <- dat_joined %>%
  filter(gender %in% c("Female/Woman", "Male/Man")) %>%
  mutate(age = case_when(age == "18-21" ~ "18-21",
                         age == "22-25" ~ "22+",
                         age == "26+" ~ "22+")) %>%
  drop_na(age)

# remove all the objects we don't need 

rm(list = c("BNFE", "dat", "dat_joined", "demo", "IUS", "STARS"))
```

## The Grammar of Graphics


```{r}
# the base
ggplot()

# adding in data set
ggplot(data = dat_final, aes(x = stars_score))

# adding layers
ggplot(data = dat_final, aes(x = stars_score)) +
  geom_histogram()

# adding more layers
ggplot(data = dat_final, aes(stars_score)) +
  geom_histogram() +
  labs(title = "My first histogram",
       x = "Mean STARS score",
       caption = "Isn't R fun")
```

## Plot types

### One categorical variable


```{r}
ggplot(dat_final, aes(x = degree_major)) +
  geom_bar()
```

### One continuous variable

```{r}
ggplot(data = dat_final, aes(stars_score)) +
  geom_histogram(binwidth = .25, 
                 boundary = 0, 
                 colour = "black",
                 fill = "lightblue") 
```

### One Continuous, One Categorical Variable

#### Grouped histograms

```{r}
# grouped by colour
ggplot(dat_final, aes(x = stars_score, fill = gender)) +
  geom_histogram(binwidth = .25, 
                 boundary = 0)

# dodged bars
ggplot(dat_final, aes(x = stars_score, fill = gender)) +
  geom_histogram(binwidth = .25, 
                 boundary = 0,
                 position = "dodge")

# faceted histogram
ggplot(dat_final, aes(x = stars_score)) +
  geom_histogram(binwidth = .25, 
                 boundary = 0) +
  facet_wrap(~gender, nrow = 2, scales = "free_y")
```

####  Density plots 

```{r}
ggplot(data = dat_final, aes(x = stars_score, fill = gender)) +
  geom_density(alpha = .4)
```

#### Boxplots

```{r}
ggplot(dat_final, aes(x = gender, y = stars_score)) +
  geom_boxplot()
```

#### Violin plots

```{r}
ggplot(dat_final, aes(x = gender, y = stars_score)) +
  geom_violin()
```

#### Violin-boxplots

The order of layers matters!

```{r}
ggplot(dat_final, aes(x = gender, y = stars_score)) +
  geom_violin() +
  geom_boxplot(width = .2)
```


### Two Continuous Variables

#### Scatterplots

```{r}

# simple scatterplot
ggplot(dat_final, aes(x = stars_score, y = ius_score)) +
  geom_point()

# add line of best fit
ggplot(dat_final, aes(x = stars_score, y = ius_score)) +
  geom_point() +
  geom_smooth(method = "lm")

# add in a grouping variable to all geoms
ggplot(dat_final, aes(x = stars_score, y = ius_score, colour = gender)) +
  geom_point() +
  geom_smooth(method = "lm")

# add in a grouping variable to one geom
ggplot(dat_final, aes(x = stars_score, y = ius_score)) +
  geom_point() +
  geom_smooth(aes(colour = gender), method = "lm")
```

## Making it look nice

### Titles and labels


```{r}
ggplot(dat_final, aes(x = stars_score, y = ius_score)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(title = "Relationship between IUS and STARS",
       subtitle = "Positive correlation",
       x = "Statistics Anxiety Score",
       y = "Intolerance of Uncertainity Score")

```

### Adding a theme



```{r}

# theme_minimal from ggplot
ggplot(dat_final, aes(x = stars_score, y = ius_score)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(title = "Relationship between IUS and STARS",
       subtitle = "Positive correlation",
       x = "Statistics Anxiety Score",
       y = "Intolerance of Uncertainity Score") +
  theme_minimal()

# theme_classic from ggplot
ggplot(dat_final, aes(x = stars_score, y = ius_score)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(title = "Relationship between IUS and STARS",
       subtitle = "Positive correlation",
       x = "Statistics Anxiety Score",
       y = "Intolerance of Uncertainity Score") +
  theme_classic()

# apa theme from jtools
ggplot(dat_final, aes(x = stars_score, y = ius_score)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(title = "Relationship between IUS and STARS",
       subtitle = "Positive correlation",
       x = "Statistics Anxiety Score",
       y = "Intolerance of Uncertainity Score") +
  theme_apa()
```

### Adding colour


```{r}
# add colour, remove legend
ggplot(dat_final, aes(x = gender, y = stars_score, fill = gender)) +
  geom_boxplot() +
  guides(fill = "none")

# use a colour blind friendly palette

ggplot(dat_final, aes(x = gender, y = stars_score, fill = gender)) +
  geom_boxplot() +
  guides(fill = "none") +
  scale_fill_brewer(palette = "Dark2")

# specify colour manually

ggplot(dat_final, aes(x = gender, y = stars_score, fill = gender)) +
  geom_boxplot() +
  guides(fill = "none") +
  scale_fill_manual(values = c("darkgrey", "lightblue"))

```

### Editing the scale


```{r}

# change the breaks
ggplot(data = dat_final, aes(stars_score)) +
  geom_histogram(binwidth = .25, 
                 boundary = 0, 
                 colour = "black",
                 fill = "lightblue") +
  scale_x_continuous(breaks = seq(from = 1, to = 5, by = 0.5))

# changing the labels - be careful of order!
ggplot(dat_final, aes(x = gender, y = stars_score, fill = gender)) +
  geom_boxplot() +
  guides(fill = "none") +
  scale_fill_manual(values = c("darkgrey", "lightblue")) +
  scale_x_discrete(labels = c("Woman", "Man"))

# changing the limits
ggplot(data = dat_final, aes(stars_score)) +
  geom_histogram(binwidth = .25, 
                 boundary = 0, 
                 colour = "black",
                 fill = "lightblue") +
  scale_x_continuous(limits = c(0,6))



```


## Going a bit wild

```{r}
# changing the labels - be careful of order!
ggplot(dat_final, aes(x = gender, y = stars_score, fill = gender)) +
  geom_jitter(width = .25, height = .2, alpha = .2) +
  geom_violin(alpha =.4) +
  geom_boxplot() +
  stat_summary(fun = "mean", geom = "point", size =2) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = .2)+
  guides(fill = "none") +
  scale_fill_manual(values = c("darkgrey", "lightblue")) +
  scale_x_discrete(labels = c("Woman", "Man"))+
  labs(title = "Gender differences in statistics anxiety",
       x = "Gender",
       y = "Statistics Anxiety Score") +
  theme_apa()+
  annotate(geom = "text",
           label = "Maybe we need DEI after all",
           x = 1.5, y = 3.5,
           hjust = 0, vjust = 1, 
           color = "black", fontface = "bold",
           angle = 45) 
```


```{r}
# marginal histograms
p <- ggplot(dat_final, aes(x = stars_score, y = ius_score)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(title = "Relationship between IUS and STARS",
       subtitle = "Positive correlation",
       x = "Statistics Anxiety Score",
       y = "Intolerance of Uncertainity Score") +
  theme_apa() 

ggMarginal(p, type = "histogram")
```

